---
title: "Data Generation"
author: "Yuchen Li (li215), section 1UG"
date: "April 14, 2019"
output: 
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Team Powers

## Standard Probability Distributions

### Normal

```{r}
genNormalPowers <- function(n, mean=0, sd=1) {
  # INPUT:
  # n is the number of teams
  
  # OUTPUT:
  # returns a vector of team powers, sorted in decreasing order
  powers <- rnorm(n, mean, sd)
  return(sort(abs(powers), decreasing=TRUE))
}
```

## Accept-Reject

```{r}
# Reference: adapted from Yuchen Li (li215), HW2, Exercise 4

acceptReject <- function(nsim, f, min, max, M) {
  # INPUT:
  # nsim is the number of simulations
  # f is the target distribution
  # min is the min value in the domain of f
  # max is the max value in the domain of f
  # max
  # M >= sup{f(x)}
  
  # OUTPUT:
  # returns a vector of random variates sampled from f, using the 
  #   Accept-Reject method with Unif(min, max) as the reference distribution
  k1 = 0             # counter for accepted samples
  j1 = 0             # number of iterations required to get desired sample size
  y1 = numeric(nsim)    # storing the sample
  while(k1 < nsim){
      u = runif(1)
      x = runif(1, min, max)  # random variate from reference distribution
      g1 = 1
      if (u < f(x) / M / g1) {
        # condition of accepting x in our sample
        k1 = k1 + 1
        y1[k1] = x
      }
      j1 = j1 + 1
  }
  return(sort(y1, decreasing=TRUE))
}
```

## Inverse CDF

```{r}
inverseCDF <- function(n, inv_cdf) {
  # INPUT:
  # n is the number of simulations
  # inv_cdf is the inverse CDF function for f
  
  # OUTPUT:
  # returns a vector of random variates sampled from PDF f,
  #   using the Inverse CDF method
  u = runif(n)
  y = numeric(n)
  for (i in 1:n) {
    y[i] = inv_cdf(u[i])
  }
  return(sort(y, decreasing=TRUE))
}
```

# Cross-Team Winning Probabilities

```{r}
genCrossTeamWinningProbabilities <- function(powers) {
  # INPUT:
  # powers is the teams powers
  
  # OUTPUT:
  # returns an n x n matrix M where M_{ij} is the probability of team-i beating team-j
  n = length(powers)
  probs = matrix(nrow=n, ncol=n)
  for (i in 1:n) {
    for (j in 1:n) {
      probs[i,j] = powers[i] / (powers[i] + powers[j])
    }
  }
  return(probs)
}
```

# Seeding

What are the other good methods than random selection? (In the data generation part, we do not have actual competition data yet.)

```{r}
# Example
sample(1:8, size=2)
```

# Examples

## Normal(0, 1) team powers
```{r}
genCrossTeamWinningProbabilities(
 genNormalPowers(4) 
)
```

## Normal(4, 10) team powers
Note the probabilities are closer to 0.5 
```{r}
genCrossTeamWinningProbabilities(
 genNormalPowers(4, mean=10, sd=0.1) 
)
```

## Beta(3, 4) team powers using Accept-Reject
```{r}
# Test `acceptReject`
f <- function(x) {return(x^2 * (1-x)^3)}
hist(rbeta(1000, 3, 4))
hist(acceptReject(1000, f, 0, 1, 0.0346))
# Actual
genCrossTeamWinningProbabilities(
 acceptReject(4, f, 0, 1, 0.0346)
)
```

## Exp(1) team powers using Inverse CDF
```{r}
# Test `inverseCDF`
inv_cdf <- function(x) {return(-log(x))}
hist(rexp(1000))
hist(inverseCDF(
    1000, 
    inv_cdf
  )
)
# Actual
genCrossTeamWinningProbabilities(
 inverseCDF(
    4, 
    inv_cdf
  )
)
```