---
title: "Bayesian Analysis"
author: 'Philip Dohm, netID: pdohm2'
date: "April 30, 2019"
output: 
    html_document:
        toc: yes
---

# Helper Functions

## Generation Functions

```{r}
# generation functions


genNormalPowers <- function(n, mean=0, sd=1) {
  # INPUT:
  # n is the number of teams
  
  # OUTPUT:
  # returns a vector of team powers, sorted in decreasing order
  powers <- rnorm(n, mean, sd)
  return(sort(abs(powers), decreasing=TRUE))
}


acceptReject <- function(nsim, f, min, max, M) {
  # INPUT:
  # nsim is the number of simulations
  # f is the target distribution
  # min is the min value in the domain of f
  # max is the max value in the domain of f
  # max
  # M >= sup{f(x)}
  
  # OUTPUT:
  # returns a vector of random variates sampled from f, using the 
  #   Accept-Reject method with Unif(min, max) as the reference distribution
  k1 = 0             # counter for accepted samples
  j1 = 0             # number of iterations required to get desired sample size
  y1 = numeric(nsim)    # storing the sample
  while(k1 < nsim){
      u = runif(1)
      x = runif(1, min, max)  # random variate from reference distribution
      g1 = 1
      if (u < f(x) / M / g1) {
        # condition of accepting x in our sample
        k1 = k1 + 1
        y1[k1] = x
      }
      j1 = j1 + 1
  }
  return(sort(y1, decreasing=TRUE))
}


inverseCDF <- function(n, inv_cdf) {
  # INPUT:
  # n is the number of simulations
  # inv_cdf is the inverse CDF function for f
  
  # OUTPUT:
  # returns a vector of random variates sampled from PDF f,
  #   using the Inverse CDF method
  u = runif(n)
  y = numeric(n)
  for (i in 1:n) {
    y[i] = inv_cdf(u[i])
  }
  return(sort(y, decreasing=TRUE))
}


genCrossTeamWinningProbabilities <- function(powers) {
  # INPUT:
  # powers is the teams powers
  
  # OUTPUT:
  # returns an n x n matrix M where M_{ij} is the probability of team-i beating team-j
  n = length(powers)
  probs = matrix(nrow=n, ncol=n)
  for (i in 1:n) {
    for (j in 1:n) {
      probs[i,j] = powers[i] / (powers[i] + powers[j])
    }
  }
  return(probs)
}
```

## Seeding Functions

```{r}
# seeding functions


#' Creates a traditional tournament style seeding arrangement. 
#'
#' @param size is the number of seeds to generate. This needs
#' to be a power of two.
#'
tournament_seeding = function(size) {
    
    #' Generates the next round of seeding based off the passed in round.
    #'
    #' @param seeds is the previous round of seeding to generate for.
    #'
    nextRound = function(seeds) {
        
        # Next round has twice the size
        nextSize = length(seeds) * 2
        out = numeric(nextSize)
        idx = 1
        
        # Make sure to match up best seed with worst
        for (seed in seeds) {
            out[idx] = seed
            out[idx + 1] = (nextSize + 1) - seed
            idx = idx + 2
        }
        
        return(out)
    }
    
    # Stop if not a power of two
    if (sum(as.numeric(intToBits(size))) != 1) {
        stop("Size must be a power of two")
    }
    
    # Special case when size is 2
    if (size == 2) {
        return(c(1, 2))
    }
    
    # Calculate number of rounds
    rounds = log2(size) - 1
    
    # Start with first two seeds
    seeds = c(1, 2)
    
    # Iteratively generate next seeds
    for (i in 1:rounds) {
        seeds = nextRound(seeds)
    }
    
    return(seeds)
}

#' Creates a random seeding arrangement. 
#'
#' @param size is the number of seeds to generate. This needs
#' to be a power of two.
#'
random_seeding = function(size) {
    # Stop if not a power of two
    if (sum(as.numeric(intToBits(size))) != 1) {
        stop("Size must be a power of two")
    }
    
    return(sample(1:size, size))
}
```

## Simulation Functions

```{r}
# simulation functions


simGame <- function(teams, matchups) {
  # function to simulate a game
  
  # INPUT:
  # teams is a vector of two teams (indices) that are "playing'
  # matchups is the matchup/probability matrix
  
  # OUTPUT:
  # returns winner of game
  
  team1 <- teams[1]
  team2 <- teams[2]
  matchup <- matchups[team1, team2]
  
  if (runif(1) < matchup) {
    return(team1)
  } else {
    return(team2)
  }
}

simRound <- function(bracket, matchups) {
  # function to simulate a round of a tournament given a bracket and probability matrix
  
  # INPUT:
  # bracket is the bracket structure
  # matchups is the matchup/probability matrix
  
  # OUTPUT:
  # returns winners of the round
  
  nGames <- length(bracket)/2
  outBracket <- c()
  for (game in 1:nGames) {
    teams <- bracket[(2*game - 1):(2*game)]
    winner <- simGame(teams, matchups)
    outBracket <- c(outBracket, winner)
  }
  
  return(outBracket)
}

# want to capture number of wins each team has per tournament
# list with the winning team
# "winner = team #
# "num wins" = c(how many wins each team has gotten)

simTournament <- function(bracket, matchups) {
  # function to simulate a tournament given a bracket and probability matrix
  
  # INPUT:
  # bracket is the bracket structure
  # matchups is the matchup/probability matrix
  
  # OUTPUT:
  # returns winner of tournament
  
  nTeams <- length(bracket)
  newBracket <- bracket
  nRounds <- log2(nTeams)
  
  results = c("Winner"=-1,"WinsPerTeam"=list(rep(0, ncol(matchups))))
  
  for (round in 1:nRounds) {
    #winsPrev = results["WinsPerTeam"]
    newBracket <- simRound(newBracket, matchups)
    for(i in newBracket){
      results$WinsPerTeam[i] = results$WinsPerTeam[i]+1
    }
  }
  
  results["Winner"] = newBracket
  return(results)
}
```

## Bayesian Functions

```{r}
# bayesian functions


# get win distribution of a bracket (via simulation)
getWinDistribution <- function(bracket, matchups, nSim=10000) {
    nTeams <- ncol(matchups)
    out <- data.frame(teams=1:nTeams, freq=numeric(nTeams))
    for(i in 1:nSim) {
        winner <- simTournament(bracket, matchups)$Winner
        out[winner, 2] <- out[winner, 2] + 1
    }
    out$teams <- as.numeric(out$teams)
    out$prob <- out$freq / nSim
    return(out)
}


# get prior and posterior win distributions of an old and new bracket
getPriorAndPosteriorWins <- function(oldBracket, newBracket, matchups, nSim=10000) {
    prior <- getWinDistribution(oldBracket, matchups, nSim)
    posterior <- getWinDistribution(newBracket, matchups, nSim)
    
    out <- prior
    colnames(out) <- c("teams", "priorFreq", "priorProb")
    out$postFreq <- posterior$freq
    out$postProb <- posterior$prob
    return(out)
}


getExpectedWins <- function(bracket, matchups, nSim=10000) {
    simResults <- replicate(nSim, simTournament(bracket, matchups), simplify='array')
    
    winsPerTeam <- c()
    for(i in 1:nSim) winsPerTeam = rbind(winsPerTeam, simResults[,i]$WinsPerTeam)
    expWins <- apply(winsPerTeam, MARGIN=2, FUN=mean)
    sdWins <- apply(winsPerTeam, MARGIN=2, FUN=sd)
    
    teams <- 1:ncol(matchups)
    out <- data.frame(teams, expWins, sdWins)
    out$teams <- as.numeric(out$teams)
    return(out)
}


getPriorAndPosteriorExpectedWins <- function(oldBracket, newBracket, matchups, nSim=10000) {
    prior <- getExpectedWins(oldBracket, matchups, nSim)
    posterior <- getExpectedWins(newBracket, matchups, nSim)
    
    out <- prior
    colnames(out) <- c("teams", "priorExp", "priorSD")
    
    totTeams <- ncol(matchups)
    oldTeams <- length(oldBracket)
    newTeams <- length(newBracket)
    
    oldRounds <- log2(totTeams) - log2(oldTeams)
    newRounds <- log2(totTeams) - log2(newTeams)
    
    out$priorExp <- out$priorExp
    out$postExp <- posterior$expWins
    out$postSD <- posterior$sdWins
    
    for (team in 1:totTeams) {
        if (any(newBracket == team)) out$postExp[team] = out$postExp[team] + newRounds
        if (any(oldBracket == team)) out$priorExp[team] = out$priorExp[team] + oldRounds
    }
    
    return(out)
}

bayesianTeam <- function(oldBracket, newBracket, matchups, nSim=10000) {
    nTeams <- ncol(matchups)
    diffRounds <- log2(length(oldBracket) / length(newBracket))
    nGames <- log2(nTeams)
    priorWPT <- postWPT <- matrix(0, nrow=nTeams, ncol=nSim)
    for (i in 1:nSim) {
        priorWPT[,i] <- simTournament(oldBracket, matchups)$WinsPerTeam
        postWPT[,i] <- simTournament(newBracket, matchups)$WinsPerTeam
        for (team in oldBracket) {
            if (any(newBracket == team)) {
                postWPT[team, i] <- postWPT[team, i] + diffRounds
            }
        }
    }
    
    out <- list()
    for (team in 1:nTeams) {
        ap <- matrix(0, nrow=nGames+1, ncol=3)
        ap <- as.data.frame(ap)
        colnames(ap) <- c("wins", "priorFreq", "postFreq")
        ap$wins <- 0:nGames
        for (win in 0:nGames) {
            ap[win+1, 2] <- length(which(priorWPT[team,] == win))
            ap[win+1, 3] <- length(which(postWPT[team,] == win))
        }
        ap$priorProb <- ap$priorFreq / nSim
        ap$postProb <- ap$postFreq / nSim
        out[[team]] <- ap
    }
    
    return(out)
}

bayesianTeamPlots <- function(listPriorPost) {
    prior <- listPriorPost[[1]]
    post <- listPriorPost[[2]]
    nTeams <- nrow(prior)
    nSim <- ncol(prior)
    
    teamPlots <- function(performance, nSim) {
        barplot(table(performance)/nSim)
    }
    
    for (i in 1:nTeams) {
        teamPlots(prior[i,], nSim)
        teamPlots(post[i,], nSim)
    }
}
```


# Analysis

We want to consider

- Prior vs. posterior when favorite win first round (compare tournament vs. random seeding)
- Prior vs. posterior when underdogs win first round (compare tournament vs. random seeding)
- Prior vs. posterior when high seeds (e.g. 1, 2, 3, or 4) are upset (compare tournament vs. random seeding)

We will look at the distribution of a teams number of wins as well as the distribution of who wins the tournament. We will use the same number of teams and the same matchup matrix (generated from Normal Powers) for all simulations and analysis.


```{r}
n <- 8 # number of teams
nRounds <- log2(n)
set.seed(1) # for consistency
M <- genCrossTeamWinningProbabilities(genNormalPowers(n)) # matchups
t <- tournament_seeding(n) # tournament seeding


# what if favorite wins every game
tFav <- t[which(t <= n/2)]
bFav <- bayesianTeam(t, tFav, M)
titles <- paste("Prior vs. Posterior for Team", 1:16)
par(mfrow=c(2,2))
for (i in 1:8) {
    df <- bFav[[i]]
    plot(postProb~wins, data=df, type='l', xlim=c(0,nRounds), ylim=c(0,1), col=i, lty=1,
         xlab="Wins", ylab="Probability", main=titles[i])
    lines(priorProb~wins, data=df, col=i, lty=2)
}

# what if underdog wins every game
tUD <- t[which(t > 8)]
bUpset <- bayesianTeam(t, tUD, M)
titles <- paste("Prior vs. Posterior for Team", 1:16)
par(mfrow=c(2,2))
for (i in 9:16) {
    df <- bUpset[[i]]
    plot(postProb~wins, data=df, type='l', xlim=c(0,nRounds), ylim=c(0,1), col=i, lty=1,
         xlab="Wins", ylab="Probability", main=titles[i])
    lines(priorProb~wins, data=df, col=i, lty=2)
}

# what if top seed is upset
t1L <- t
t1L[1] <- 16
b1L <- bayesianTeam(t, t1L, M)
titles <- paste("Prior vs. Posterior for Team", 1:16)
par(mfrow=c(2,2))
for (i in 1:16) {
    df <- b1L[[i]]
    plot(postProb~wins, data=df, type='l', xlim=c(0,nRounds), ylim=c(0,1), col=i, lty=1,
         xlab="Wins", ylab="Probability", main=titles[i])
    lines(priorProb~wins, data=df, col=i, lty=2)
}

t2L <- t
t2L[9] <- 15
b1L <- bayesianTeam(t, t2L, M)
plot(NULL, xlim=c(0,4), ylim=c(0,1), xlab="wins", ylab="probability")
for (i in 1:8) {
    df <- b1L[[i]]
    lines(postProb~wins, data=df, col=i, lty=1)
    lines(priorProb~wins, data=df, col=i, lty=2)
}
```

