---
title: "Results"
date: "Due May 3, 2019"
output:
  html_document:
    theme: readable
    toc: yes
  pdf_document:
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.pos = '!h')
```

## Data Generation Functions

```{r}
genNormalPowers <- function(n, mean=0, sd=1) {
  # INPUT:
  # n is the number of teams
  
  # OUTPUT:
  # returns a vector of team powers, sorted in decreasing order
  powers <- rnorm(n, mean, sd)
  return(sort(abs(powers), decreasing=TRUE))
}

# Reference: adapted from Yuchen Li (li215), HW2, Exercise 4

acceptReject <- function(nsim, f, min, max, M) {
  # INPUT:
  # nsim is the number of simulations
  # f is the target distribution
  # min is the min value in the domain of f
  # max is the max value in the domain of f
  # max
  # M >= sup{f(x)}
  
  # OUTPUT:
  # returns a vector of random variates sampled from f, using the 
  #   Accept-Reject method with Unif(min, max) as the reference distribution
  k1 = 0             # counter for accepted samples
  j1 = 0             # number of iterations required to get desired sample size
  y1 = numeric(nsim)    # storing the sample
  while(k1 < nsim){
      u = runif(1)
      x = runif(1, min, max)  # random variate from reference distribution
      g1 = 1
      if (u < f(x) / M / g1) {
        # condition of accepting x in our sample
        k1 = k1 + 1
        y1[k1] = x
      }
      j1 = j1 + 1
  }
  return(sort(y1, decreasing=TRUE))
}

inverseCDF <- function(n, inv_cdf) {
  # INPUT:
  # n is the number of simulations
  # inv_cdf is the inverse CDF function for f
  
  # OUTPUT:
  # returns a vector of random variates sampled from PDF f,
  #   using the Inverse CDF method
  u = runif(n)
  y = numeric(n)
  for (i in 1:n) {
    y[i] = inv_cdf(u[i])
  }
  return(sort(y, decreasing=TRUE))
}

genCrossTeamWinningProbabilities <- function(powers) {
  # INPUT:
  # powers is the teams powers
  
  # OUTPUT:
  # returns an n x n matrix M where M_{ij} is the probability of team-i beating team-j
  n = length(powers)
  probs = matrix(nrow=n, ncol=n)
  for (i in 1:n) {
    for (j in 1:n) {
      probs[i,j] = powers[i] / (powers[i] + powers[j])
    }
  }
  return(probs)
}
```

## Seeding Functions

```{r}
#' Creates a traditional tournament style seeding arrangement. 
#'
#' @param size is the number of seeds to generate. This needs
#' to be a power of two.
#'
tournament_seeding = function(size) {
    
    #' Generates the next round of seeding based off the passed in round.
    #'
    #' @param seeds is the previous round of seeding to generate for.
    #'
    nextRound = function(seeds) {
        
        # Next round has twice the size
        nextSize = length(seeds) * 2
        out = numeric(nextSize)
        idx = 1
        
        # Make sure to match up best seed with worst
        for (seed in seeds) {
            out[idx] = seed
            out[idx + 1] = (nextSize + 1) - seed
            idx = idx + 2
        }
        
        return(out)
    }
    
    # Stop if not a power of two
    if (sum(as.numeric(intToBits(size))) != 1) {
        stop("Size must be a power of two")
    }
    
    # Special case when size is 2
    if (size == 2) {
        return(c(1, 2))
    }
    
    # Calculate number of rounds
    rounds = log2(size) - 1
    
    # Start with first two seeds
    seeds = c(1, 2)
    
    # Iteratively generate next seeds
    for (i in 1:rounds) {
        seeds = nextRound(seeds)
    }
    
    return(seeds)
}

#' Creates a seeding arrangement that is left skewed.
#'
#' @param size is the number of seeds to generate. This needs
#' to be a power of two.
#'
left_skewed_seeding = function(size) {
    
    # Stop if not a power of two
    if (sum(as.numeric(intToBits(size))) != 1) {
        stop("Size must be a power of two")
    }
    
    # Special case when size is 2
    if (size == 2) {
        return(c(1, 2))
    }
    
    # Create seeds and left/right branches
    seeds = 1:size
    left_seeds = numeric(size / 2)
    right_seeds = numeric(size / 2)
    
    # Indexes for the left/right branches
    left_idx = 1
    right_idx = 1
    
    # Keep track of which branch will get the next seed
    left_turn = TRUE
    
    # For each iteration add the first and last seed to the
    # current branch
    for (i in 1:(size / 2)) {
        if (left_turn) {
            left_seeds[left_idx] = seeds[i]
            left_seeds[left_idx + 1] = seeds[(size + 1) - i]
            left_idx = left_idx + 2
            left_turn = FALSE
        } else {
            right_seeds[right_idx] = seeds[i]
            right_seeds[right_idx + 1] = seeds[(size + 1) - i]
            right_idx = right_idx + 2
            left_turn = TRUE
        }
    }
    
    # Combine branches
    c(left_seeds, right_seeds)
}

#' Creates a random seeding arrangement. 
#'
#' @param size is the number of seeds to generate. This needs
#' to be a power of two.
#'
random_seeding = function(size) {
    # Stop if not a power of two
    if (sum(as.numeric(intToBits(size))) != 1) {
        stop("Size must be a power of two")
    }
    
    return(sample(1:size, size))
}
```

## Simulation Functions

```{r}
simGame <- function(teams, matchups) {
  # function to simulate a game
  
  # INPUT:
  # teams is a vector of two teams (indices) that are "playing'
  # matchups is the matchup/probability matrix
  
  # OUTPUT:
  # returns winner of game
  
  team1 <- teams[1]
  team2 <- teams[2]
  matchup <- matchups[team1, team2]
  
  if (runif(1) < matchup) {
    return(team1)
  } else {
    return(team2)
  }
}

simRound <- function(bracket, matchups) {
  # function to simulate a round of a tournament given a bracket and probability matrix
  
  # INPUT:
  # bracket is the bracket structure
  # matchups is the matchup/probability matrix
  
  # OUTPUT:
  # returns winners of the round
  
  nGames <- length(bracket)/2
  outBracket <- c()
  for (game in 1:nGames) {
    teams <- bracket[(2*game - 1):(2*game)]
    winner <- simGame(teams, matchups)
    outBracket <- c(outBracket, winner)
  }
  
  return(outBracket)
}

# want to capture number of wins each team has per tournament
# list with the winning team
# "winner = team #
# "num wins" = c(how many wins each team has gotten)

simTournament <- function(bracket, matchups) {
  # function to simulate a tournament given a bracket and probability matrix
  
  # INPUT:
  # bracket is the bracket structure
  # matchups is the matchup/probability matrix
  
  # OUTPUT:
  # returns winner of tournament
  
  nTeams <- length(bracket)
  newBracket <- bracket
  nRounds <- log2(nTeams)
  
  results = c("Winner"=-1,"WinsPerTeam"=list(rep(0, ncol(matchups))))
  
  for (round in 1:nRounds) {
    #winsPrev = results["WinsPerTeam"]
    newBracket <- simRound(newBracket, matchups)
    for(i in newBracket){
      results$WinsPerTeam[i] = results$WinsPerTeam[i]+1
    }
  }
  
  results["Winner"] = newBracket
  return(results)
}
```

## Resampling Functions

```{r}
#function to generate bootstrap replicates
#INPUT:
#x is vector of team ranking/ other measure of 
#team power in different simulations
#B is number of replicates to generate
#estimator is the statistic we want, set as mean by
#default
#l is the length of bootstrap resample in each run
#set to length of x by default
#OUTPUT
#a list of bootstrap replicates, bias ans standard error
boot_replicate <- function(x, B = 5000, estimator = mean, l = length(x)){
  boots <- replicate(B, estimator(sample(x,l,replace=TRUE)))
  sd_b <- sd(boots)
  bias_b <- mean(boots - estimator(x))
  list(boot = boots, bias = bias_b, se = sd_b)
}

#function to generate jackknife replicates
#INPUT:
#x is vector of team ranking/ other measure of 
#team power in different simulations
#estimator is the statistic we want, set as mean by
#default
#OUTPUT
#a list of jackknife replicates, bias ans standard error
jack_replicate <- function(x, estimator = mean){
  n <- length(x)
  jack <- numeric(n)
  #perform jackknife
  for(i in 1:n)
    jack[i] <- estimator(x[-i])
  #bias using jackknife
  bias_j <- (n-1)*(mean(jack)-estimator(x))
  sumsq=sum((jack-mean(jack))^2)
  se_j <- sqrt((n-1)/n)*sqrt(sumsq)
  list(jack = jack,bias = bias_j, se = se_j)
}
```

## Results

```{r}
getWinners = function(results) {
    winnerList = c()
    for(i in 1:(length(results) / 2)){
        winnerList = c(winnerList,results[,i]$Winner)
    }
    return(winnerList)
}

getWinsPerTeam = function(results) {
    winsPerTeamList = c()
    for(i in 1:(length(results) / 2)){
        winsPerTeamList = rbind(winsPerTeamList,results[,i]$WinsPerTeam)
    }
    return(winsPerTeamList)
}
```

# Additional functions

```{r}
# Re-generate Team Powers and Cross-Team Winning Probabilities before each simulation
simTournamentRegenerateSeedsAndProbs <- function(genSeeds, genProbs) {
    bracket <- genSeeds()
    matchups <- genProbs()
    return(simTournament(bracket, matchups))
}
```

# How do the parameters affect the probability that the strongest team wins?

## $\mu$ in Normal

```{r}
numTeams = 8
nSim = 1000

# Normal(1, 1) team powers
genSeeds <- function() {
    return(random_seeding(numTeams))
}
genProbs <- function() {
    return(genCrossTeamWinningProbabilities(genNormalPowers(numTeams, mean=1)))
}

norm_11_results = replicate(
    nSim, 
    simTournamentRegenerateSeedsAndProbs(genSeeds, genProbs), 
    simplify="array"
)
norm_11_winners = getWinners(norm_11_results)
norm_11_boot_res <- boot_replicate(
    norm_11_winners,
    estimator=function(x) mean(x==1)
)

# Normal(3, 1) team powers
genSeeds <- function() {
    return(random_seeding(numTeams))
}
genProbs <- function() {
    return(genCrossTeamWinningProbabilities(genNormalPowers(numTeams, mean=3)))
}

norm_31_results = replicate(
    nSim, 
    simTournamentRegenerateSeedsAndProbs(genSeeds, genProbs), 
    simplify="array"
)
norm_31_winners = getWinners(norm_31_results)
norm_31_boot_res <- boot_replicate(
    norm_31_winners,
    estimator=function(x) mean(x==1)
)

# Normal(9, 1) team powers
genSeeds <- function() {
    return(random_seeding(numTeams))
}
genProbs <- function() {
    return(genCrossTeamWinningProbabilities(genNormalPowers(numTeams, mean=9)))
}

norm_91_results = replicate(
    nSim, 
    simTournamentRegenerateSeedsAndProbs(genSeeds, genProbs), 
    simplify="array"
)
norm_91_winners = getWinners(norm_91_results)
norm_91_boot_res <- boot_replicate(
    norm_91_winners,
    estimator=function(x) mean(x==1)
)

# Box plots
boxplot(
    norm_11_boot_res$boot,
    norm_31_boot_res$boot,
    norm_91_boot_res$boot,
    names=c("Normal(1,1)", "Normal(3,1)", "Normal(9,1)")
)
title("Probability that the strongest team wins")
```

**Interpretation:** Controlling the standard deviation, as the mean increases, the team powers are more similar to each other. Hence the probability that the strongest team wins would decrease.

## $\sigma$ in Normal

```{r}
numTeams = 8
nSim = 1000

# Normal(9, 1) team powers
genSeeds <- function() {
    return(random_seeding(numTeams))
}
genProbs <- function() {
    return(genCrossTeamWinningProbabilities(genNormalPowers(numTeams, mean=9, sd=1)))
}

norm_91_results = replicate(
    nSim, 
    simTournamentRegenerateSeedsAndProbs(genSeeds, genProbs), 
    simplify="array"
)
norm_91_winners = getWinners(norm_91_results)
norm_91_boot_res <- boot_replicate(
    norm_91_winners,
    estimator=function(x) mean(x==1)
)

# Normal(9, 3) team powers
genSeeds <- function() {
    return(random_seeding(numTeams))
}
genProbs <- function() {
    return(genCrossTeamWinningProbabilities(genNormalPowers(numTeams, mean=9, sd=3)))
}

norm_93_results = replicate(
    nSim, 
    simTournamentRegenerateSeedsAndProbs(genSeeds, genProbs), 
    simplify="array"
)
norm_93_winners = getWinners(norm_93_results)
norm_93_boot_res <- boot_replicate(
    norm_93_winners,
    estimator=function(x) mean(x==1)
)

# Normal(9, 5) team powers
genSeeds <- function() {
    return(random_seeding(numTeams))
}
genProbs <- function() {
    return(genCrossTeamWinningProbabilities(genNormalPowers(numTeams, mean=9, sd=5)))
}

norm_95_results = replicate(
    nSim, 
    simTournamentRegenerateSeedsAndProbs(genSeeds, genProbs), 
    simplify="array"
)
norm_95_winners = getWinners(norm_95_results)
norm_95_boot_res <- boot_replicate(
    norm_95_winners,
    estimator=function(x) mean(x==1)
)

# Box plots
boxplot(
    norm_91_boot_res$boot,
    norm_93_boot_res$boot,
    norm_95_boot_res$boot,
    names=c("Normal(9,1)", "Normal(9,3)", "Normal(9,5)")
)
title("Probability that the strongest team wins")
```

**Interpretation:** Controlling the mean, as the standard deviation increases, the team powers are more different from each other. Hence the probability that the strongest team wins would increase.

## $\alpha$ in Gamma($\alpha$, $\beta$)

```{r}
numTeams = 8
nSim = 1000

gammaPDF <- function(x, a, b) {
    return(b^a / gamma(a) * x^(a-1) * exp(-b*x))
}

# Gamma(2,2) team powers
f <- function(x) {
    return(gammaPDF(x, 2, 2))
}
genSeeds <- function() {
    return(random_seeding(numTeams))
}
genProbs <- function() {
    return(genCrossTeamWinningProbabilities(
        genCrossTeamWinningProbabilities(
         acceptReject(numTeams, f, 0, 25, 0.2)
        )
    ))
}

gamma_22_results = replicate(
    nSim, 
    simTournamentRegenerateSeedsAndProbs(genSeeds, genProbs), 
    simplify="array"
)
gamma_22_winners = getWinners(gamma_22_results)
gamma_22_boot_res <- boot_replicate(
    gamma_22_winners,
    estimator=function(x) mean(x==1)
)

# Gamma(3,2) team powers
f <- function(x) {
    return(gammaPDF(x, 3, 2))
}
genSeeds <- function() {
    return(random_seeding(numTeams))
}
genProbs <- function() {
    return(genCrossTeamWinningProbabilities(
        genCrossTeamWinningProbabilities(
         acceptReject(numTeams, f, 0, 25, 0.2)
        )
    ))
}

gamma_32_results = replicate(
    nSim, 
    simTournamentRegenerateSeedsAndProbs(genSeeds, genProbs), 
    simplify="array"
)
gamma_32_winners = getWinners(gamma_32_results)
gamma_32_boot_res <- boot_replicate(
    gamma_32_winners,
    estimator=function(x) mean(x==1)
)

# Gamma(4,2) team powers
f <- function(x) {
    return(gammaPDF(x, 4, 2))
}
genSeeds <- function() {
    return(random_seeding(numTeams))
}
genProbs <- function() {
    return(genCrossTeamWinningProbabilities(
        genCrossTeamWinningProbabilities(
         acceptReject(numTeams, f, 0, 25, 0.2)
        )
    ))
}

gamma_42_results = replicate(
    nSim, 
    simTournamentRegenerateSeedsAndProbs(genSeeds, genProbs), 
    simplify="array"
)
gamma_42_winners = getWinners(gamma_42_results)
gamma_42_boot_res <- boot_replicate(
    gamma_42_winners,
    estimator=function(x) mean(x==1)
)

# Box plots
boxplot(
    gamma_22_boot_res$boot,
    gamma_32_boot_res$boot,
    gamma_42_boot_res$boot,
    names=c("Gamma(2,2)", "Gamma(3,2)", "Gamma(4,2)")
)
title("Probability that the strongest team wins")
```

**Interpretation:** Controlling the rate parameter, as the shape parameter increases, the team powers are less right-skewed. Hence the probability that the strongest team wins would decrease.

## $\lambda$ in Exponential($\lambda$)

```{r}
numTeams = 8
nSim = 1000

# Exponential(1) team powers
inv_cdf <- function(x) {return(-log(x) / 1)}
genSeeds <- function() {
    return(random_seeding(numTeams))
}
genProbs <- function() {
    return(genCrossTeamWinningProbabilities(
        genCrossTeamWinningProbabilities(
         inverseCDF(
            numTeams, 
            inv_cdf
          )
        )
    ))
}

exp_1_results = replicate(
    nSim, 
    simTournamentRegenerateSeedsAndProbs(genSeeds, genProbs), 
    simplify="array"
)
exp_1_winners = getWinners(exp_1_results)
exp_1_boot_res <- boot_replicate(
    exp_1_winners,
    estimator=function(x) mean(x==1)
)

# Exponential(2) team powers
inv_cdf <- function(x) {return(-log(x) / 2)}
genSeeds <- function() {
    return(random_seeding(numTeams))
}
genProbs <- function() {
    return(genCrossTeamWinningProbabilities(
        genCrossTeamWinningProbabilities(
         inverseCDF(
            numTeams, 
            inv_cdf
          )
        )
    ))
}

exp_2_results = replicate(
    nSim, 
    simTournamentRegenerateSeedsAndProbs(genSeeds, genProbs), 
    simplify="array"
)
exp_2_winners = getWinners(exp_2_results)
exp_2_boot_res <- boot_replicate(
    exp_2_winners,
    estimator=function(x) mean(x==1)
)

# Exponential(3) team powers
inv_cdf <- function(x) {return(-log(x) / 1)}
genSeeds <- function() {
    return(random_seeding(numTeams))
}
genProbs <- function() {
    return(genCrossTeamWinningProbabilities(
        genCrossTeamWinningProbabilities(
         inverseCDF(
            numTeams, 
            inv_cdf
          )
        )
    ))
}

exp_3_results = replicate(
    nSim, 
    simTournamentRegenerateSeedsAndProbs(genSeeds, genProbs), 
    simplify="array"
)
exp_3_winners = getWinners(exp_3_results)
exp_3_boot_res <- boot_replicate(
    exp_3_winners,
    estimator=function(x) mean(x==1)
)

# Box plots
boxplot(
    exp_1_boot_res$boot,
    exp_2_boot_res$boot,
    exp_3_boot_res$boot,
    names=c("Exponential(1)", "Exponential(2)", "Exponential(3)")
)
title("Probability that the strongest team wins")
```

**Interpretation:** As the rate parameter increases, the team powers are more right-skewed. Hence the probability that the strongest team wins would increase.
